---
title: "Module 1: Tabular Data"
subtitle: "Working with larger-than-RAM data using duckdbfs"
author: "Mei Collins"
format: 
  html:
    embed-resources: true
---

## Case Study: Global Supply Chains

We will be working with [EXIOBASE 3.8.1](https://source.coop/youssef-harby/exiobase-3), a global Multi-Regional Input-Output (MRIO) database. This dataset tracks economic transactions between sectors and regions, along with their environmental impacts (emissions, resource use, etc.).

**Data description:**
- **Coverage**: 44 countries + 5 rest-of-world regions.
- **Timeframe**: 1995–2022.
- **Content**: Economic transactions (Z matrix), final demand (Y matrix), and environmental stressors (F matrix).
- **Format**: Cloud-optimized Parquet, partitioned by year and matrix type.

## Setup

```{r}
library(duckdbfs)
library(dplyr)
library(ggplot2)
```

## Connecting to data and quick exploration

```{r}
# Open the dataset - filtering for F_satellite matrix (environmental stressors)

# Note: The F matrix is called "F_satellite" in this dataset
# Using wildcard pattern to capture all years (1995-2022)
matrix_f <- open_dataset("s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/year=*/format=ixi/matrix=F_satellite")

# Check the structure
matrix_f

# Preview the first few rows
matrix_f |> 
  head(10) |> 
  collect()

# See what years are available
matrix_f |> 
  distinct(year) |> 
  arrange(year) |>
  collect()
```

I want to identify data on the CO2 production country by country, over time. 

Which stressors are related to CO2?
```{r}
# Filter stressors to find those related to CO2
co2_stressors <- matrix_f |>
  distinct(stressor) |>
  filter(grepl("CO2|carbon dioxide", stressor, ignore.case = TRUE)) |>
  collect()

co2_stressors
```

Great, now we know there are 6 stressors. Let's identify which regions are the top 5 CO2 emitters based on these stressors, and then we can explore their emissions over time.

```{r}
# Filter data for CO2 stressors and aggregate by region
co2_by_region <- matrix_f |>
  filter(stressor %in% co2_stressors$stressor) |>
  group_by(region) |>
  summarise(total_co2 = sum(value, na.rm = TRUE)) |>
  arrange(desc(total_co2)) |>
  head(5) |>
  collect()

co2_by_region

top5_regions <- co2_by_region$region
```

Now that we know the top 5 emitting regions, we can plot their CO2 emissions over time. 

First let's pull the relevant data as a time series
```{r}
# Filter and aggregate CO2 emissions by region and year for top 5
co2_time_series <- matrix_f |>
  filter(stressor %in% co2_stressors$stressor,
         region %in% top5_regions) |>
  group_by(region, year) |>
  summarise(total_co2 = sum(value, na.rm = TRUE), .groups = "drop") |>
  collect()

```

Then let's make a graph and save it
```{r}
# Plot emissions over time
emissions_graph <- ggplot(co2_time_series, aes(x = year, y = total_co2, color = region)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "CO2 Emissions Over Time - Top 5 Emitting Regions",
       x = "Year",
       y = "Total CO2 Emissions",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Save the plot to a file
ggsave("top5emitters_timeseries.png", plot = emissions_graph, width = 10, height = 6)

```

Other questions we can explore with this data include:
- which countries are decreasing in emissions and why?
- what are the top sectors and how do they change over time? 


Let's tackle the first question: which countries are decreasing in emissions and why?

```{r}
# Calculate the change in emissions from 1995 to 2022 for each region
emissions_change <-  matrix_f|>
  filter(stressor %in% co2_stressors$stressor) |>
  group_by(region) |>
  summarise(emissions_1995 = sum(value[year == 1995]),
            emissions_2022 = sum(value[year == 2022]),
  )|>
    mutate(percent_change = (emissions_2022 - emissions_1995) / emissions_1995 * 100) |>
  arrange(percent_change) |>
  head(5) |>
  collect()     

emissions_change
```


Now plot the emissions over time for these five regions

```{r}
# Get the top 5 regions with the largest decrease in emissions
top5_decreasing_regions <- emissions_change$region


# Filter and aggregate CO2 emissions by region and year for top 5 decreasing regions
co2_decreasing_time_series <- matrix_f |>      
    filter(stressor %in% co2_stressors$stressor,
             region %in% top5_decreasing_regions) |>
    group_by(region, year) |>
    summarise(total_co2 = sum(value, na.rm = TRUE), .groups = "drop") |>
    collect()   

# Plot emissions over time for decreasing regions
decreasing_emissions_graph <- ggplot(co2_decreasing_time_series, aes(x = year, y = total_co2, color = region)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "CO2 Emissions Over Time - Top 5 Decreasing Regions",
       x = "Year",
       y = "Total CO2 Emissions",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom") 

# Save the plot to a file
ggsave("top5decreasing_emitters_timeseries.png", plot = decreasing_emissions_graph, width = 10, height = 6)
```
