---
title: "Module 1: Tabular Data"
subtitle: "Working with larger-than-RAM data using duckdbfs"
author: "ESPM 288"
format: html
---

## Introduction

In this module, we will explore high-performance workflows for tabular data. We will use `duckdbfs` to work with datasets that are larger than available RAM by leveraging DuckDB's streaming and remote file access capabilities.

## Case Study: Global Supply Chains

We will be working with [EXIOBASE 3.8.1](https://source.coop/youssef-harby/exiobase-3), a global Multi-Regional Input-Output (MRIO) database. This dataset tracks economic transactions between sectors and regions, along with their environmental impacts (emissions, resource use, etc.).

**Data description:**
- **Coverage**: 44 countries + 5 rest-of-world regions.
- **Timeframe**: 1995–2022.
- **Content**: Economic transactions (Z matrix), final demand (Y matrix), and environmental stressors (F matrix).
- **Format**: Cloud-optimized Parquet, partitioned by year and matrix type.

## Setup

```{r}
library(duckdbfs)
library(dplyr)

```

## Exercise 1: connecting to remote data

We can open the entire dataset without downloading it using `open_dataset()`. The data is hosted on Source Cooperative. The `**` pattern allows recursive scanning of the partitioned parquet files.

```{r}
# Open the dataset - filtering for F_satellite matrix (environmental stressors)

# Note: The F matrix is called "F_satellite" in this dataset
# Using wildcard pattern to capture all years (1995-2022)
matrix_f <- open_dataset("s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/year=*/format=ixi/matrix=F_satellite")

# Check the structure
matrix_f

# Preview the first few rows
matrix_f |> 
  head(10) |> 
  collect()

# See what years are available
matrix_f |> 
  distinct(year) |> 
  arrange(year) |>
  collect()
```


```{r}

# Explore the column names and types in the matrix 

# See the column names and types in matrix_f
matrix_f |> 
  head(1) |> 
  collect() |>
  glimpse()

# Or just get column names
matrix_f |> 
  head(1) |> 
  collect() |>
  names()
```


## Exercise 2: filtering and aggregating data
Now that we have the dataset open, let's perform some filtering and aggregation operations.

```{r}
# Aggregate emissions by country and produce a graph over all years
# for the top 5 emitting countries


#find the top 5 emitting countries
top_countries <- matrix_f |> 
  group_by(region) |> 
  summarise(total_emissions = sum(value, na.rm = TRUE)) |> 
  arrange(desc(total_emissions)) |> 
  head(5) |> 
  collect()

print(top_countries)

top_countries <- c('CN','US', 'IN', "WM", 'WA')

# produce graph of emissions over time for top 5 countries
emissions_over_time <- matrix_f |> 
  filter(region %in% top_countries) |> 
  group_by(year, region) |> 
  summarise(annual_emissions = sum(value, na.rm = TRUE)) |> 
  collect()

#plot the results
library(ggplot2)
emissions_graph <- ggplot(emissions_over_time, aes(x = year, y = annual_emissions, color = region)) +
  geom_line(size = 1) +
  labs(title = "Annual Emissions by Top 5 Emitting Countries",
       x = "Year",
       y = "Annual Emissions") +
  theme_minimal()

# Save the plot to a file
ggsave("emissions_plot.png", plot = emissions_graph, width = 10, height = 6)

# Print message
print("Plot saved to emissions_plot.png")
```